---
title: "Peer Assessment 1"
author: "Bob Carasik"
date: "Saturday, May 09, 2015"
output: 
  html_document:
    keep_md: TRUE
---

In the first R code chunk, I load libraries, set the directory,  load the dataset,
and clean the data by preparing a dataset with null data omitted:

## Loading and preprocessing the data

```{r, loadingdata, echo=FALSE, cache=TRUE}
require(ggplot2)
library(ggplot2)
library(grid)
library(gridExtra)
setwd("C:/Users/Bob Carasik/Documents/Bob/ReproducibleResearch/Project1")
if(!exists("rawdata")) {
  rawdata <- read.csv("activity.csv")
}  
  #  clean the data by omitting measurements where steps are "NA"
actdata <- rawdata[complete.cases(rawdata[,c(1)]),]
```

## What is mean total number of steps taken per day?
Here, I create the plot for the first question and create a histogram of steps per day.  I count the number of days for each range of total steps taken. I then report the mean and median of the total number of steps taken per day.

```{r, stepsperday, echo=FALSE, cache=TRUE}
thedays <-  unique (actdata$date)
daycount <- length(thedays)
daytotals <- data.frame(Date = thedays,
                        MeanSteps = double(daycount), 
                        TotalSteps = integer(daycount)
)
for(i in 1:daycount) {
  day <- thedays[i]
  daytotals[i,]$MeanSteps <- mean(actdata[actdata$date==day,]$steps)
  daytotals[i,]$TotalSteps <- sum(actdata[actdata$date==day,]$steps)
} 
meanStepsPerDay <- mean(daytotals$TotalSteps)
medianStepsPerDay <- median(daytotals$TotalSteps)
hplot <- hist(daytotals$TotalSteps, breaks = 8)
# print (hplot )
cat(" mean steps per day from raw data ", meanStepsPerDay, " median steps per day ", medianStepsPerDay, "\n")
```


## What is the average daily activity pattern?

The second part of the assignment is the average daily activity pattern
I created a time series plot of the 5-minute intervals on the x-axis and the average number of steps taken, averaged across all days on the y-axis.

I then identify the 5-minute interval that contains the maximum number of steps on average across all the days in the dataset. 
```{r, averageactivity, echo=FALSE}
intervals <-  unique (actdata$interval)
intervalcount <- length(intervals)
intervaltotals <- data.frame(Interval = intervals,
                        MeanSteps = double(intervalcount), 
                        TotalSteps = integer(intervalcount),
                        WeekdaySteps = integer(intervalcount),
                        WeekendSteps = integer(intervalcount))
for(i in 1:intervalcount) {
  interval <- intervals[i]
  intervaltotals[i,]$MeanSteps <- mean(actdata[actdata$interval==interval,]$steps)
  intervaltotals[i,]$TotalSteps <- sum(actdata[actdata$interval==interval,]$steps)
}
maxintervalStepCount <- max(intervaltotals$TotalSteps)
maxStepsInterval <- intervaltotals[which(intervaltotals$TotalSteps == 
                       max(intervaltotals$TotalSteps)),]$Interval
# Mean steps per interval, for all days
with ( intervaltotals, plot(Interval, MeanSteps))
cat(" The interval with the most steps is ", maxStepsInterval, "\n")
```

## Imputing missing values

I next create a copy of the raw data set with values imputed where the steps data is absent.  I simply assign the mean number of steps for a given interval when there is no steps data for an interval for a day in the dataset.  I then re-calculate mean and median steps per day where the missing values have been imputed

```{r, imputingdata, echo=FALSE}
 imputeddata <- rawdata
 rrows <- nrow(rawdata)
 for (i in 1:rrows) {
  if (is.na(rawdata[i,]$steps )) {
    imputeddata[i,]$steps <- 
      as.integer(intervaltotals[ intervaltotals$Interval == 
                                   rawdata[i,]$interval,]$MeanSteps)
  }
}
for(i in 1:daycount) {
  day <- thedays[i]
  daytotals[i,]$MeanSteps <- mean(imputeddata[imputeddata$date==day,]$steps)
  daytotals[i,]$TotalSteps <- sum(imputeddata[imputeddata$date==day,]$steps)
} 
meanStepsPerDayImputed <- mean(daytotals$TotalSteps)
medianStepsPerDayImputed <- median(daytotals$TotalSteps)

hplot <- hist(daytotals$TotalSteps, breaks = 8, main="total steps by day from imputed data")
cat(" mean steps per day from imputed data ", meanStepsPerDayImputed, " median steps per day ", medianStepsPerDayImputed, "\n")
```
## Are there differences in activity patterns between weekdays and weekends?

Now I add a "weekdays" factor to the imputed data, and I make two plots for steps by interval for weekends and weekdays

```{r, weekendfactor, echo=FALSE}
library(ggplot2)
library(grid)
library(gridExtra)
imputeddata$weekend <- 
  ifelse( weekdays(as.Date(imputeddata$date)) == "Saturday" | 
          weekdays(as.Date(imputeddata$date)) == "Sunday", "Yes", "No")
#
for(i in 1:intervalcount) {
  interval <- intervals[i]
  intervaltotals[i,]$WeekdaySteps <- 
      sum(imputeddata[imputeddata$interval==interval & 
                        imputeddata$weekend == "No",]$steps)
  intervaltotals[i,]$WeekendSteps <- 
      sum(imputeddata[imputeddata$interval==interval & 
                        imputeddata$weekend == "Yes",]$steps)
}
weekendplot <- ggplot( intervaltotals, aes(x = Interval, y = WeekendSteps))    + geom_point() + ggtitle("Weekend steps")
weekdayplot <- ggplot(intervaltotals, aes(x = Interval, y = WeekendSteps))     + geom_point() + ggtitle("Weekday steps")
grid.arrange(weekendplot, weekdayplot, ncol=2)
```
 